---
title: "past_data"
date: "2023-04-21"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = F, warning = F, error = F, fig.align = "center", collapse = T)
options(knitr.kable.NA="", knitr.table.format = "html")
```

```{r libraries}
library(raster)
library(sf)
library(dplyr)
library(ggplot2)

library(spdep)
```

# Extract past data

```{r wr}
wr.sf <- readRDS("wr.sf.rds")

```

## HILDA

HILDA data are downloaded from https://www.wur.nl/en/research-results/chair-groups/environmental-sciences/laboratory-of-geo-information-science-and-remote-sensing/models/hilda/hilda-data-downloads.html

Hilda categories:

* 77: Water
* 66: Other land
* 55: Grass/schrubland
* 45: Forest, mixed
* 44: Forest, deciduous broad leaf
* 43: Forest: deciduous needle
* 42: Forest, evergreen borad leaf
* 41: Forest, evergreen needle
* 40: Forest, other
* 33: Pasture
* 22: Cropland
* 11: Urban

```{r extract-data-hilda-1899}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1899 <- raster::extract(hilda.raster, layer = 1, wr.sf)

saveRDS(hilda.1899, "hilda.1899.wr.rds")

hilda.tab <- sapply(hilda.1899, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1899.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1899.rds")
```

```{r extract-data-hilda-1909}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1909 <- raster::extract(hilda.raster, layer = 2, wr.sf)

saveRDS(hilda.1909, "hilda.1909.wr.rds")

hilda.tab <- sapply(hilda.1909, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1909.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1909.rds")
```

```{r extract-data-hilda-1919}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1919 <- raster::extract(hilda.raster, layer = 3, wr.sf)

saveRDS(hilda.1919, "hilda.1919.wr.rds")

hilda.tab <- sapply(hilda.1919, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1919.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1919.rds")
```

```{r extract-data-hilda-1929}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1929 <- raster::extract(hilda.raster, layer = 4, wr.sf)

saveRDS(hilda.1929, "hilda.1929.wr.rds")

hilda.tab <- sapply(hilda.1929, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1929.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1929.rds")
```

```{r extract-data-hilda-1939}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1939 <- raster::extract(hilda.raster, layer = 5, wr.sf)

saveRDS(hilda.1939, "hilda.1939.wr.rds")

hilda.tab <- sapply(hilda.1939, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1939.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1939.rds")
```

```{r extract-data-hilda-1949}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1949 <- raster::extract(hilda.raster, layer = 6, wr.sf)

saveRDS(hilda.1949, "hilda.1949.wr.rds")

hilda.tab <- sapply(hilda.1949, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1949.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1949.rds")
```

```{r extract-data-hilda-1959}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1959 <- raster::extract(hilda.raster, layer = 7, wr.sf)

saveRDS(hilda.1959, "hilda.1959.wr.rds")

hilda.tab <- sapply(hilda.1959, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1959.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1959.rds")
```

```{r extract-data-hilda-1969}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1969 <- raster::extract(hilda.raster, layer = 8, wr.sf)

saveRDS(hilda.1969, "hilda.1969.wr.rds")

hilda.tab <- sapply(hilda.1969, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1969.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1969.rds")
```

```{r extract-data-hilda-1979}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1979 <- raster::extract(hilda.raster, layer = 9, wr.sf)

saveRDS(hilda.1979, "hilda.1979.wr.rds")

hilda.tab <- sapply(hilda.1979, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1979.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1979.rds")
```

```{r extract-data-hilda-1989}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1989 <- raster::extract(hilda.raster, layer = 10, wr.sf)

saveRDS(hilda.1989, "hilda.1989.wr.rds")

hilda.tab <- sapply(hilda.1989, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1989.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1989.rds")
```

```{r extract-data-hilda-1999}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.1999 <- raster::extract(hilda.raster, layer = 11, wr.sf)

saveRDS(hilda.1999, "hilda.1999.wr.rds")

hilda.tab <- sapply(hilda.1999, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.1999.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.1999.rds")
```

```{r extract-data-hilda-2009}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.2009 <- raster::extract(hilda.raster, layer = 12, wr.sf)

saveRDS(hilda.2009, "hilda.2009.wr.rds")

hilda.tab <- sapply(hilda.2009, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.2009.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.2009.rds")
```

```{r extract-data-hilda-2019}
hilda.raster <- "HILDA_plus_decadal.nc" %>% raster()

hilda.2019 <- raster::extract(hilda.raster, layer = 13, wr.sf)

saveRDS(hilda.2019, "hilda.2019.wr.rds")

hilda.tab <- sapply(hilda.2019, function(x) tabulate(x,77))
hilda.tab.area <- hilda.tab*prod(res(hilda.raster))
catchment.area <- colSums(hilda.tab.area)
hilda.tab.prop <- sweep(hilda.tab.area,2,catchment.area, FUN = "/")
names(hilda.tab.prop) <- wr.sf$gid
saveRDS(hilda.tab.prop,"hilda.2019.tab.prop.Rdata")

forest <- colSums(hilda.tab.prop[c(40,41,42,43,44,45),]) %>% as.data.frame() %>% setNames("forest") %>% tibble::rownames_to_column(var = "gid")
saveRDS(forest,"forest.2019.rds")
```

## Sdep

 Annual mean deposition in kg/m2/s, averaged in the water drainage basins polygons.
 
```{r load-wr}
wr.sf <- readRDS("wr.sf.rds")
```

```{r extract-sulfur-1899}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(589:600))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(589:600))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(589:600))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(589:600))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1899.rds")
```

```{r extract-sulfur-1909}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(709:720))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(709:720))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(709:720))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(709:720))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1909.rds")
```

```{r extract-sulfur-1919}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(829:840))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(829:840))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(829:840))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(829:840))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1919.rds")
```

```{r extract-sulfur-1929}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(949:960))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(949:960))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(949:960))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(949:960))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1929.rds")
```

```{r extract-sulfur-1939}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1069:1080))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1069:1080))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1069:1080))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1069:1080))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1939.rds")
```

```{r extract-sulfur-1949}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1189:1200))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1189:1200))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1189:1200))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1189:1200))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1949.rds")
```

```{r extract-sulfur-1959}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1309:1320))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1309:1320))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1309:1320))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1309:1320))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1959.rds")
```

```{r extract-sulfur-1969}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1429:1440))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1429:1440))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1429:1440))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1429:1440))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1969.rds")
```

```{r extract-sulfur-1979}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1979.rds")
```

```{r extract-sulfur-1979}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1549:1560))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1979.rds")
```

```{r extract-sulfur-1989}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1669:1680))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1669:1680))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1669:1680))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1669:1680))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1989.rds")
```

```{r extract-sulfur-1999}
dryso2 <- raster("Sdep/dryso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1789:1800))
dryso4 <- raster("Sdep/dryso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1789:1800))
wetso2 <- raster("Sdep/wetso2_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1789:1800))
wetso4 <- raster("Sdep/wetso4_aero_NorESM1-ME_historical_r1i1p1_185001-200512.nc") %>% stack(bands = c(1789:1800))

dryso2.df <- extract(dryso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso2.df) <- c("gid","area","dso2")
dryso4.df <- extract(dryso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(dryso4.df) <- c("gid","area","dso4")
wetso2.df <- extract(wetso2, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso2.df) <- c("gid","area","wso2")
wetso4.df <- extract(wetso4, wr.sf, sp = T, df = T, na.rm = T, fun = mean)
names(wetso4.df) <- c("gid","area","wso4")

sdep.df <- merge(dryso2.df, dryso4.df, by = "gid") %>% merge(wetso2.df, by = "gid") %>% merge(wetso4.df, by = c("gid","area"))
sdep.df$tsdep <- rowSums(sdep.df@data[,c(3:6)])

saveRDS(sdep.df, "sdep.df.1999.rds")
```

Does not exist for 2009. EMEP?

# Predict TOC

```{r load-model}
toc_model <- readRDS("sem.fen.log.rds")
wr.sf <- readRDS("wr.sf.rds")

lims <- c(0,20)

```

```{r wr.kmat, eval = F}
wr.spdf <- SpatialPointsDataFrame(st_coordinates(st_centroid(wr.sf)), st_drop_geometry(wr.sf))
wr.kmat <- knearneigh(wr.spdf, k = 100) %>% knn2nb() %>% nb2listw()
saveRDS(wr.kmat, "wr.kmat.rds")
```

```{r predict-1899, eval = F}
forest <- readRDS("forest.1899.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1899.rds")
runoff <- readRDS("runoff.1899.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1899 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1899$logRunoff <- wr.sf.1899$runoff %>% log10()
wr.sf.1899$TSdep <- wr.sf.1899$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1899$logTSdep <- log10(wr.sf.1899$TSdep)

wr.sf.1899$logTOC.fit <- predict(toc_model, newdata = wr.sf.1899, wr.kmat)[,1]
wr.sf.1899$TOC.fit <- 10^wr.sf.1899$logTOC.fit
wr.sf.1899$TOC.export <- with(wr.sf.1899, TOC.fit*runoff* 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1899, "wr.sf.1899.rds")
```

```{r map-toc-1899, eval = F}
map.toc.1899 <- ggplot(wr.sf.1899)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1899", fill = "TOC concentration\nin mg/L in 1899")+
  theme_void()

ggsave("map.toc.1899.png", map.toc.1899)
```

```{r predict-1909, eval = F}
forest <- readRDS("forest.1909.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1909.rds")
runoff <- readRDS("runoff.1909.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1909 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1909$logRunoff <- wr.sf.1909$runoff %>% log10()
wr.sf.1909$TSdep <- wr.sf.1909$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1909$logTSdep <- log10(wr.sf.1909$TSdep)

wr.sf.1909$logTOC.fit <- predict(toc_model, newdata = wr.sf.1909, wr.kmat)[,1]
wr.sf.1909$TOC.fit <- 10^wr.sf.1909$logTOC.fit
wr.sf.1909$TOC.export <- with(wr.sf.1909, TOC.fit*runoff* 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1909, "wr.sf.1909.rds")
```

```{r map-toc-1909, eval = F}
map.toc.1909 <- ggplot(wr.sf.1909)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1909", fill = "TOC concentration\nin mg/L in 1909")+
  theme_void()

ggsave("map.toc.1909.png", map.toc.1909)
```

```{r predict-1919, eval = F}
forest <- readRDS("forest.1919.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1919.rds")
runoff <- readRDS("runoff.1919.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1919 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1919$logRunoff <- wr.sf.1919$runoff %>% log10()
wr.sf.1919$TSdep <- wr.sf.1919$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1919$logTSdep <- log10(wr.sf.1919$TSdep)

wr.sf.1919$logTOC.fit <- predict(toc_model, newdata = wr.sf.1919, wr.kmat)[,1]
wr.sf.1919$TOC.fit <- 10^wr.sf.1919$logTOC.fit
wr.sf.1919$TOC.export <- with(wr.sf.1919, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1919, "wr.sf.1919.rds")
```

```{r map-toc-1919, eval = F}
map.toc.1919 <- ggplot(wr.sf.1919)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1919", fill = "TOC concentration\nin mg/L in 1919")+
  theme_void()

ggsave("map.toc.1919.png", map.toc.1919)
```

```{r predict-1929, eval = F}
forest <- readRDS("forest.1929.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1929.rds")
runoff <- readRDS("runoff.1929.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1929 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1929$logRunoff <- wr.sf.1929$runoff %>% log10()
wr.sf.1929$TSdep <- wr.sf.1929$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1929$logTSdep <- log10(wr.sf.1929$TSdep)

wr.sf.1929$logTOC.fit <- predict(toc_model, newdata = wr.sf.1929, wr.kmat)[,1]
wr.sf.1929$TOC.fit <- 10^wr.sf.1929$logTOC.fit
wr.sf.1929$TOC.export <- with(wr.sf.1929, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1929, "wr.sf.1929.rds")
```

```{r map-toc-1929, eval = F}
map.toc.1929 <- ggplot(wr.sf.1929)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1929", fill = "TOC concentration\nin mg/L in 1929")+
  theme_void()

ggsave("map.toc.1929.png", map.toc.1929)
```

```{r predict-1939, eval = F}
forest <- readRDS("forest.1939.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1939.rds")
runoff <- readRDS("runoff.1939.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1939 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1939$logRunoff <- wr.sf.1939$runoff %>% log10()
wr.sf.1939$TSdep <- wr.sf.1939$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1939$logTSdep <- log10(wr.sf.1939$TSdep)

wr.sf.1939$logTOC.fit <- predict(toc_model, newdata = wr.sf.1939, wr.kmat)[,1]
wr.sf.1939$TOC.fit <- 10^wr.sf.1939$logTOC.fit
wr.sf.1939$TOC.export <- with(wr.sf.1939, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1939, "wr.sf.1939.rds")
```

```{r map-toc-1939, eval = F}
map.toc.1939 <- ggplot(wr.sf.1939)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1939", fill = "TOC concentration\nin mg/L in 1939")+
  theme_void()

ggsave("map.toc.1939.png", map.toc.1939)
```

```{r predict-1949, eval = F}
forest <- readRDS("forest.1949.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1949.rds")
runoff <- readRDS("runoff.1949.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1949 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1949$logRunoff <- wr.sf.1949$runoff %>% log10()
wr.sf.1949$TSdep <- wr.sf.1949$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1949$logTSdep <- log10(wr.sf.1949$TSdep)

wr.sf.1949$logTOC.fit <- predict(toc_model, newdata = wr.sf.1949, wr.kmat)[,1]
wr.sf.1949$TOC.fit <- 10^wr.sf.1949$logTOC.fit
wr.sf.1949$TOC.export <- with(wr.sf.1949, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1949, "wr.sf.1949.rds")
```

```{r map-toc-1949, eval = F}
map.toc.1949 <- ggplot(wr.sf.1949)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1949", fill = "TOC concentration\nin mg/L in 1949")+
  theme_void()

ggsave("map.toc.1949.png", map.toc.1949)
```

```{r predict-1959, eval = F}
forest <- readRDS("forest.1959.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1959.rds")
runoff <- readRDS("runoff.1959.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1959 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1959$logRunoff <- wr.sf.1959$runoff %>% log10()
wr.sf.1959$TSdep <- wr.sf.1959$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1959$logTSdep <- log10(wr.sf.1959$TSdep)

wr.sf.1959$logTOC.fit <- predict(toc_model, newdata = wr.sf.1959, wr.kmat)[,1]
wr.sf.1959$TOC.fit <- 10^wr.sf.1959$logTOC.fit
wr.sf.1959$TOC.export <- with(wr.sf.1959, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1959, "wr.sf.1959.rds")
```

```{r map-toc-1959, eval = F}
map.toc.1959 <- ggplot(wr.sf.1959)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1959", fill = "TOC concentration\nin mg/L in 1959")+
  theme_void()

ggsave("map.toc.1959.png", map.toc.1959)
```

```{r predict-1969, eval = F}
forest <- readRDS("forest.1969.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1969.rds")
runoff <- readRDS("runoff.1969.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1969 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1969$logRunoff <- wr.sf.1969$runoff %>% log10()
wr.sf.1969$TSdep <- wr.sf.1969$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1969$logTSdep <- log10(wr.sf.1969$TSdep)

wr.sf.1969$logTOC.fit <- predict(toc_model, newdata = wr.sf.1969, wr.kmat)[,1]
wr.sf.1969$TOC.fit <- 10^wr.sf.1969$logTOC.fit
wr.sf.1969$TOC.export <- with(wr.sf.1969, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1969, "wr.sf.1969.rds")
```

```{r map-toc-1969, eval = F}
map.toc.1969 <- ggplot(wr.sf.1969)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1969", fill = "TOC concentration\nin mg/L in 1969")+
  theme_void()

ggsave("map.toc.1969.png", map.toc.1969)
```

```{r predict-1979, eval = F}
forest <- readRDS("forest.1979.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1979.rds")
runoff <- readRDS("runoff.1979.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1979 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1979$logRunoff <- wr.sf.1979$runoff %>% log10()
wr.sf.1979$TSdep <- wr.sf.1979$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1979$logTSdep <- log10(wr.sf.1979$TSdep)

wr.sf.1979$logTOC.fit <- predict(toc_model, newdata = wr.sf.1979, wr.kmat)[,1]
wr.sf.1979$TOC.fit <- 10^wr.sf.1979$logTOC.fit
wr.sf.1979$TOC.export <- with(wr.sf.1979, TOC.fit*runoff* 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1979, "wr.sf.1979.rds")
```

```{r map-toc-1979, eval = F}
map.toc.1979 <- ggplot(wr.sf.1979)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1979", fill = "TOC concentration\nin mg/L in 1979")+
  theme_void()

ggsave("map.toc.1979.png", map.toc.1979)
```

```{r predict-1989, eval = F}
forest <- readRDS("forest.1989.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1989.rds")
runoff <- readRDS("runoff.1989.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1989 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1989$logRunoff <- wr.sf.1989$runoff %>% log10()
wr.sf.1989$TSdep <- wr.sf.1989$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1989$logTSdep <- log10(wr.sf.1989$TSdep)

wr.sf.1989$logTOC.fit <- predict(toc_model, newdata = wr.sf.1989, wr.kmat)[,1]
wr.sf.1989$TOC.fit <- 10^wr.sf.1989$logTOC.fit
wr.sf.1989$TOC.export <- with(wr.sf.1989, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1989, "wr.sf.1989.rds")
```

```{r map-toc-1989, eval = F}
map.toc.1989 <- ggplot(wr.sf.1989)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1989", fill = "TOC concentration\nin mg/L in 1989")+
  theme_void()

ggsave("map.toc.1989.png", map.toc.1989)
```

```{r predict-1999, eval = F}
forest <- readRDS("forest.1999.rds") %>% filter(!is.na(forest))
names(forest) <- c("gid","forest")

sdep <- readRDS("sdep.df.1999.rds")
runoff <- readRDS("runoff.1999.wr.rds") %>% as.data.frame() %>% filter(!is.na(runoff))

wr.sf.1999 <- merge(wr.sf, forest, by = "gid") %>% merge(sdep, by = "gid") %>% merge(runoff, by = "gid")

wr.sf.1999$logRunoff <- wr.sf.1999$runoff %>% log10()
wr.sf.1999$TSdep <- wr.sf.1999$tsdep * (-1) * 365 * 60 * 60 * 1e6 # converts to mg/year
wr.sf.1999$logTSdep <- log10(wr.sf.1999$TSdep)

wr.sf.1999$logTOC.fit <- predict(toc_model, newdata = wr.sf.1999, wr.kmat)[,1]
wr.sf.1999$TOC.fit <- 10^wr.sf.1999$logTOC.fit
wr.sf.1999$TOC.export <- with(wr.sf.1999, TOC.fit*runoff * 365 * 60 * 60 * 1e6 * area) # mg/L * kg/m2/y * m2 -> mg/year

saveRDS(wr.sf.1999, "wr.sf.1999.rds")
```

```{r map-toc-1999, eval = F}
map.toc.1999 <- ggplot(wr.sf.1999)+geom_sf(aes(fill = TOC.fit, col = TOC.fit))+
   scale_colour_continuous_divergingx(palette = "RdYlBu", aesthetics = c("col","fill"), mid = 10, rev = T, limits = lims, na.value = "black")+
  labs(col = "TOC concentration\nin mg/L in 1999", fill = "TOC concentration\nin mg/L in 1999")+
  theme_void()

ggsave("map.toc.1999.png", map.toc.1999)
```

For 2009 and 2019: Need to download data from EMEP and fit a model on NLS with EMEP data

# Compare exports

```{r create-import basins, eval = F}
wr.sf <- readRDS("wr.sf.rds") 
ggplot() + geom_sf(data = head(wr.sf), col = "black")
coords.wr <- st_coordinates(wr.sf)

basins <- st_read("basins_all_sea.shp") %>% st_cast("POLYGON")
st_crs(basins) <- st_crs(4326)
basins$sea <- with(basins, ifelse(Poly %in% c(1:161), "Baltic",
                                  ifelse(Poly %in% c(162:253), "North",
                                    "NCC")))
coords <- st_coordinates(basins)


ggplot()+ geom_sf(data = basins, aes(col = sea, fill = sea))

wr.sf.mercator <- st_transform(wr.sf, st_crs(4326))
ggplot() + geom_sf(data = head(wr.sf.mercator,100), col = "black")
wr.coords <- st_coordinates(wr.sf.mercator)

sf_use_s2(FALSE)
test <- st_join(st_centroid(wr.sf.mercator, of), basins, join = st_within)
ggplot()+geom_sf(data = test, aes(fill = sea, color = sea, size = area), na.value = "black")+
  geom_sf(data = basins, aes(col = sea), fill = NA)+
  ylim(55,71)

test2 <- merge(wr.sf.mercator, st_drop_geometry(test), by = c("gid","area"))
ggplot()+geom_sf(data = test2, aes(fill = sea, color = sea), na.value = "black")+
  geom_sf(data = basins, aes(col = sea), fill = NA)+
  ylim(55,71)
```

```{r common}
wr.sf.1899 <- readRDS("wr.sf.1899.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1899$year <- "1899"

wr.sf.1909 <- readRDS("wr.sf.1909.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1909$year <- "1909"

wr.sf.1919 <- readRDS("wr.sf.1919.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1919$year <- "1919"

wr.sf.1929 <- readRDS("wr.sf.1929.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1929$year <- "1929"

wr.sf.1939 <- readRDS("wr.sf.1939.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1939$year <- "1939"

wr.sf.1949 <- readRDS("wr.sf.1949.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1949$year <- "1949"

wr.sf.1959 <- readRDS("wr.sf.1959.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1959$year <- "1959"

wr.sf.1969 <- readRDS("wr.sf.1969.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1969$year <- "1969"

wr.sf.1979 <- readRDS("wr.sf.1979.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1979$year <- "1979"

wr.sf.1989 <- readRDS("wr.sf.1989.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1989$year <- "1989"

wr.sf.1999 <- readRDS("wr.sf.1999.rds") %>% select(c("gid","TOC.fit","TOC.export","geometry"))
wr.sf.1999$year <- "1999"

wr.sf.past <- rbind(wr.sf.1899, wr.sf.1909) %>% rbind(wr.sf.1919) %>% 
              rbind(wr.sf.1929) %>% rbind(wr.sf.1939) %>% rbind(wr.sf.1949) %>% 
              rbind(wr.sf.1959) %>% rbind(wr.sf.1969) %>% rbind(wr.sf.1979) %>% 
              rbind(wr.sf.1989) %>% rbind(wr.sf.1999)

wr.sf.past.mercator <- st_transform(wr.sf.past, st_crs(4326))
wr.sf.exp <- st_join(st_centroid(wr.sf.past.mercator),basins, join = st_within)

saveRDS(wr.sf.exp, "wr.sf.exp.rds")
```


```{r plot-exp}
basins.exp <- wr.sf.exp %>% st_drop_geometry() %>% group_by(sea,year) %>% summarise(TOCyearly = sum(TOC.export))
basins.exp$yearlyTOC.Tg <- basins.exp$TOCyearly * 1e-15 # from mg to Tg

ggplot(basins.exp)+geom_point(aes(x=year, y = yearlyTOC.Tg, col = sea))
```